Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands

### 2024-11-13 index.qmd
writing

key-points:
  - Base flow sustains dryland river ecology, but assessing its groundwater contribution faces data limitations
  - Recursive filtering and machine learning predict BFI and streamflow-associated recharge
  -

@@

The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm **(chen2016xgboost)** to predict BFI at ungauged locations using catchment characteristics as predictors. The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems. It has grown in popularity in recent years due to its rapid, efficient, and scalable nature, and has proven effective in similar environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily**; @ni2020streamflow**] and classifying land use/land cover data [@Georganos2018Very].

XGBoost uses gradient boosting on decision tree algorithms to leverage the combination of multiple very low variance models into one overall prediction. Gradient boosting works sequentially, with the first tree being learned on the known target values, and each subsequent tree being learned on the errors of the tree prior. The algorithm then applies a weighting to each tree based on this boosting to determine how much the tree will contribute towards the final output. The final prediction is obtained by the input data traversing all $n$ weighted trees in the learned ensemble. The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (Table \ref{tab-######}##}##}##}. Certain predictors were subdivided according to their areal coverage of each basin (e.g. land cover was divided into 16 subdivisions).

@@

: **Table 1.** Basin-characteristic variables used as predictors in XGBoost model.

@@

@@

#### Predictor Importance

@@

>>>>

key-points:
  - Base flow sustains dryland river ecology, but assessing its groundwater contribution faces data limitations
  - Recursive filtering and machine learning predict BFI 
  -

@@

The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors. The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems. It has grown in popularity in recent years due to its rapid, efficient, and scalable nature, and has proven effective in similar environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily**; @ni2020streamflow**] and classifying land use/land cover data [@Georganos2018Very].

XGBoost uses gradient boosting on decision tree algorithms to leverage the combination of multiple very low variance models into one overall prediction. Gradient boosting works sequentially, with the first tree being learned on the known target values, and each subsequent tree being learned on the errors of the tree prior. The algorithm then applies a weighting to each tree based on this boosting to determine how much the tree will contribute towards the final output. The final prediction is obtained by the input data traversing all $n$ weighted trees in the learned ensemble. The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (@tbl-predictors). Certain predictors were subdivided according to their areal coverage of each basin (e.g. land cover was divided into 16 subdivisions).

@@

: **Table 1.** Basin-characteristic variables used as predictors in XGBoost model. {#tbl-predictors}

### Statistical Analyses

Statisical analyses were performed on mean annual BFI values of instrumented streamgages. We examined the presence of temporal trends in mean annual BFI using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric]. The Mann-Kendall was used to determine trends in BFI and trend significance of gauged reaches during their period of record.

Trends with a $\rho \le 0.05$ are considered significant.

@@

The long-term average BFI for the 245 gauged reaches across Arizona is illustrated in **FIG BFI-INSTRUMENTED.** The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona. The highly karstic geology of this region facilitates the rapid movement of subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023]. Relatively high BFI values (\>0.8) are found at the spring-fed headwaters of the Verde River (Del Rio Spring) and the spring-fed headwaters of Fossil Creek. These results are consistent with interpolated BFI values reported by @wolock2003.

The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consisently low BFI values (\< 0.2). This is likely due to low-yielding perched aquifers underlying the Defiance Uplift in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02]. This region also typifies a pattern of seen along most major rivers in the study area: upstream reaches indicate higher BFI values while the downstream reaches display lower BFI values. This is assumed to be due to greater groundwater-surface water interaction at the headwaters of streams (likely due to spring outlets) and the dilution of base flow while moving down stream (**CITATION**). This pattern is seen along the Gila River, Verde River, and Little Colorado Rivers in Arizona.

#### Trends in BFI {#sec-bfi-trends}

@@

@@

|                              | n    | R^2^  | MSE    | RMSE  | MAE   | NSE   | pbias |
|------------------------------|------|-------|--------|-------|-------|-------|-------|
| Climate - Monsoon Dominated  | 3345 | 0.707 | 0.0166 | 0.129 | 0.077 | 0.694 | -12.6 |
| Climate - Snowmelt Dominated | 6152 | 0.723 | 0.017  | 0.131 | 0.094 | 0.715 | -3.7  |
| PhysRegion - Basin&Range     | 7754 | 0.749 | 0.018  | 0.133 | 0.089 | 0.741 | -5.9  |
| PhysRegion - CO Plateau      | 1743 | 0.806 | 0.013  | 0.115 | 0.081 | 0.801 | -4.4  |
| Climate - Warm-Wet           | 1613 | 0.690 | 0.015  | 0.124 | 0.081 | 0.679 | -9.9  |
| Climate - Warm-Dry           | 3134 | 0.758 | 0.023  | 0.151 | 0.097 | 0.745 | -8.1  |
| Climate - Cool-Wet           | 3134 | 0.710 | 0.013  | 0.115 | 0.085 | 0.708 | -2.5  |
| Climate - Cool-Dry           | 1616 | 0.809 | 0.014  | 0.118 | 0.082 | 0.803 | -5    |
| Slope - High                 | 4398 | 0.753 | 0.014  | 0.120 | 0.084 | 0.746 | -4.4  |
| Slope - Low                  | 5099 | 0.759 | 0.019  | 0.138 | 0.090 | 0.750 | -6.9  |

: Table 2. Performance of model predictions for BFI for all sites split by various classifications. n is number of sites, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sucliffe efficiency, and pbias is percent bias. {#tbl-performance}

#### Predictor Importance {#sec-predictor-importance}
